---
title: "Dataframes - In Class lab"
author: "Fatima Tanvir"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


```{r}
# In class Lab: Dash is the a womens soccer team
dash_scored<- c(1, 0, 3, 1)
dash_against<- c(5, 0, 2, 3)
opponents <- c("North Carolina Courage", "Racing Louisville FC", "Bay FC",  "Washington Spirit")

games <- data.frame(dash_scored, dash_against)
games$diff <- games$dash_against - games$dash_scored
games$won <- games$dash_scored > games$dash_against
games$opponents <- opponents

view(games)
```



# Working with loaded in csv files


Use the `read.csv()` function to read the data from the `gates_donations.csv` file (available on Canvas, make sure you save it in the project repository folder) into a variable called `grants` 

```{r}
## if you save it in the same folder as this file this should work. 
## But you may need edit the string to reflect the actual file name: 

grants <- read_csv("gates_donations.csv")

head(grants)
```

Because you're in an RProject, you don't need to set your working directory in RStudio! The working directory is automatically set to the repository location. (You can use `getwd()` to confirm this)

- Use the glimpse function to look at the loaded data

```{r}
glimpse(grants)


```


- Create a variable `org` that contains the `organization` column of the dataset

```{r}
org <- grants$organization
```


- Confirm that the "organization" column is a vector using the `is.vector()` function. (This is a useful debugging tip if you hit errors later!)

```{r}
is.vector(org)

```


Now you can ask some interesting questions about the dataset:

- What was the mean grant value?

```{r}
grant_value <- mean(grants$total_amount)
grant_value
```


- What was the dollar amount of the largest grant?

```{r}
largest_grant <- max(grants$total_amount)
largest_grant
```


- What was the dollar amount of the smallest grant?

```{r}
smallest_grant <- min(grants$total_amount)
smallest_grant
```


- Which organization received the largest grant?

```{r}
largest_grant_org <- grants$organization[which.max(grants$total_amount)]
largest_grant_org
```


- Which organization received the smallest grant?

```{r}
smallest_grant_org <- grants$organization[which.min(grants$total_amount)]
smallest_grant_org
```

- How many grants were awarded in 2010?

```{r}
grants_2010_count <- sum(grants$start_year == 2010)
grants_2010_count
```


- How much was awarded in 2010? 

```{r}
grants_2010_total <- sum(grants$total_amount[grants$start_year == 2010])
grants_2010_total
```


# Working with built-in data sets

Load R's "USPersonalExpenditure" dataset using the `data()` function

This will produce a data frame called `USPersonalExpenditure`
```{r}
data("USPersonalExpenditure")
```

The variable `USPersonalExpenditure` is now accessible to you. Unfortunately, it's not a data frame (it's actually a matrix, you can find that out using) 

Test this using the `is.data.frame()` function:

```{r}
is.data.frame(USPersonalExpenditure)
```

Luckily, you can pass the USPersonalExpenditure variable as an argument to the `data.frame()` function to convert it a data frame.

Do this, storing the result in a new variable

```{r}
us_expenditure_df <- as.data.frame(USPersonalExpenditure)

```


What are the current column names of your dataframe?
```{r}
colnames(us_expenditure_df)
```


Consider: why are they so strange? (Think about whether you could use a number like 1940 with dollar notation!)

What are the current row names of your dataframe?
```{r}
rownames(us_expenditure_df)
```


Add a column "Category" to your data frame that contains the rownames

```{r}
us_expenditure_df$Category <- rownames(us_expenditure_df)
```

Consider how this data isn't tidy. Use the `pivot_longer()` function from last class to make the data tidy. 

```{r}
tidy_us_expenditure <- pivot_longer(
  us_expenditure_df,
  cols = -Category,
  names_to = "Year",
  values_to = "Expenditure"
)
```


How much money was spent on personal care in 1940?
```{r}
personal_care_1940 <- tidy_us_expenditure$Expenditure[
  tidy_us_expenditure$Category == "Personal Care" & tidy_us_expenditure$Year == "1940"
]
personal_care_1940
```


How much money was spent on Food and Tobacco in 1960?
```{r}
food_tobacco_1960 <- tidy_us_expenditure$Expenditure[
  tidy_us_expenditure$Category == "Food and Tobacco" & tidy_us_expenditure$Year == "1960"
]
food_tobacco_1960
```


What was the highest expenditure category in 1960?
```{r}
expenditure_1960 <- tidy_us_expenditure[tidy_us_expenditure$Year == "1960", ]
highest_expenditure_1960 <- expenditure_1960$Category[which.max(expenditure_1960$Expenditure)]
highest_expenditure_1960
```


Define a function `lowest_category` that takes in a year as a parameter, and returns the lowest spending category of that year
```{r}
lowest_category <- function(year) {
  expenditure_year <- tidy_us_expenditure[tidy_us_expenditure$Year == as.character(year), ]
  lowest_category_year <- expenditure_year$Category[which.min(expenditure_year$Expenditure)]
  return(lowest_category_year)
}
```

```{r}
lowest_category(1940)
```


Using your function, determine the lowest spending category of each year
Hint: use the `sapply()` function to apply your function to a vector of years

```{r}
years <- unique(tidy_us_expenditure$Year)
lowest_categories <- sapply(years, lowest_category)
lowest_categories
```
